Laurens de Graaf

### 2015-05-06 writeup/paper.tex
More write-up done

\section{\emph{Laurencillo}} %% Introduction

@@

\section{\emph{Tigre}} %% The Spineless, Tagless G-Machine

@@

The operations are broken up by the STG machine into four possible operations:
evaluation, entering, returning a construct, and returning an integer, given
as \emph{Eval $e$ $\rho$}, \emph{Enter $a$}, \emph{ReturnCon $c$ $xs$}, and
\emph{ReturnInt $i$} in the reduction semantics. We implement each of these
as an \emph{operation node} in \lstinline{op.py}, wherein each has a
\lstinline{step} method that expects a configuration using that operator
and dispatches appropriately, returning the new machine configuration.

\section{\emph{Francesca}} %% RPython

In this section, we discuss a number of intricacies that require careful
consideration during implementation of the STG machine.

\subsection{Unique Stack Implementations}\label{sec:typeinf}

As previously state, RPython attempts exhaustive type inference for structures
during compilation. This approach has forced us into using three different
stack implementations, all identical in code; we would like to use a single
stack defined as \lstinline{data/stack.py}, but these authors could not find
a way to circumvent the type inference to allow such reuse.
Indeed, we would suggest that this is one of the largest
hurdles in constructing a project such as this in RPython: there is a non-zero
amount of code duplication required to simply exist in the RPython ecosystem.

Furthermore, the later sections of \citet{spj:stgmachine} describe an
optimization for combining stacks that would suggest that we can combine these
stacks by carefully packaging each different node in a new object to indicate
which of the original stacks it lived on\footnote{There are other mechanisms to
avoid this discussed in the original work, but they require subtle bit
manipulation that would need to be faked in a higher-level implementation.}. 
Unfortunately, this would add one further level of indirection for the compiler
and further obfuscate our implementation, and it is unclear if we could gain
speed for the effort.  Further experimentation should be done, here, however.

\subsection{The Funny Case of Laziness}

While standard lazy languages provide forcing at primitive operations, the
STG machine departs from this explicit approach. Instead, a primitive operation
assumes that its arguments are pre-evaluated, and this became troublesome when
testing them. To further elaborate this point, observe a standard definition
of addition in the STG language:
\begin{code}[language=Haskell]
+ = \ {} \\n {n,m} -> case n of
                        n' -> case m of
                                m' -> #+ n' m'
\end{code}
This snippet is interpreted as: 
\begin{center}
\emph{
``Plus is a lambda with no free variables that
is should not be updated. When applied, it will \lstinline{case} on
\lstinline{n} and then \lstinline{m}, using the forcing position of
\lstinline{case} to ensure that both inputs are evaluated before using the
primitive addition operator \lstinline{\#+} on the result.''
}
\end{center}
While not particularly tied to the implementation in RPython, we found this
insight invaluable during implementation and testing and so we have included
it here. This manifests with the following dispatch for addition in our
implementation:
\begin{code}
elif expr_type is ast.ast.PrimOp:
  if cexp.oper == "+":
    lookups = vals(code.env,
                   global_env, 
                   cexp.atoms)
    x1      = lookups[0]
    x2      = lookups[1]
    assert isinstance(x1,ast.ast.ValAST)
    assert isinstance(x2,ast.ast.ValAST)
    res     = x1.value + x2.value
    config.code = op.ReturnInt(res) 
\end{code}
We first ensure that we are performing addition and then look up the operators
with \lstinline{vals} (which uses the local and global environments for variable
lookup). Next, we ensure that each of the resulting lookups are valid Value
nodes, extract their values, and then construct a \lstinline{ReturnIn} operation
code for the configuration.

\section{\emph{Neptune} n\'{e}e \emph{Fortune}} %% Conclusion / Future Work

>>>>

\section{\emph{Laurencillo}~-- Introduction} %% Introduction

@@

\section{\emph{Tigre}~-- The STG Machine} %% The Spineless, Tagless G-Machine

@@

The operations are broken up by the STG machine into four possible operations:
evaluation, entering, returning a construct, and returning an integer, given
as \emph{Eval $e$ $\rho$}, \emph{Enter $a$}, \emph{ReturnInt $i$}, and
\emph{ReturnCon $c$ $xs$} in the reduction semantics. We implement each of these
as an \emph{operation node} in \lstinline{op.py}, wherein each has a
\lstinline{step} method that expects a configuration using that operator and
dispatches appropriately, returning the new machine configuration.

\subsection{Operations}

There are four operations in the STG machine which indicate the next
step the machine will take. We implement each of these as a \lstinline{step}
attached to an operation node in \lstinline{op.py}, allowing us to easily
modularize the main loop.

\paragraph{\textit{Eval e $\rho$}.}
The \textit{Eval} step evaluates an operation, yielding either a 
sub-expression to evaluate (in the case of a \lstinline{case} operation)
or an \textit{Enter} operation with a heap address location (such as when
evaluating an application).
This is, by far, the most grotesque part of the implementation: to implement the
evaluation step, we extract the expression node and branch on its type,
constructing a new configuration as necessary based on the result.

\paragraph{\textit{Enter a}.}
The \textit{Enter} step enters a heap address, which is a \emph{closure} stored
in the heap. This operation also handles thunk updates and partial applications,
constructing an updated thunk in the former case and a new thunk with additional
input in the latter.

\paragraph{\textit{ReturnInt i}.}
This operation indicates that an expression has evaluated to an integer $i$. The
return stack will contain a \lstinline{case} expression to match against this
result. We present this case in full in Figure~\ref{fig:ReturnInt}; the 
\lstinline{int_case_lookup} operator takes the relevant cases and identifies the
appropriate one. We also indicate a \emph{default case}, which uses a binder
with a default variable to bind the result if it does not match an integer
literal.

\begin{figure}
\begin{code}
def step(self,config):
    debug("Int return")

    if config.ret_stack.empty():
      raise Exception('...')
    else: 
      retk      = config.ret_stack.pop()
      if type(retk) is ast.cont.CaseCont:
        ret_env   = retk.env.copy()
        ret_alts  = retk.alts
        value     = self.value

        default,default_var,body = 
          self.int_case_lookup(value, ret_alts)

        if default:
          if (default_var is not None):
            var = default_var
            ret_env[var] = 
              ast.ast.Value(value, True)

        config.code = op.Eval(body, ret_env)

    return config

  def int_case_lookup(self,val,ret_alts):
    # Assuming everything is a LitAlt
    for alt in ret_alts.alternates: 
      if alt.literal == val:
        return (False, None, alt.rhs)

    return (True, 
            ret_alts.default.binder, 
            ret_alts.default.rhs)
\end{code}
\caption{\label{fig:returnint}The \textit{ReturnInt} operation.}
\end{figure}

\paragraph{\textit{ReturnCon c xs}.}
This operation indicates that an expression has evaluated to a constructor $c$
with arguments in the $xs$. The return stack (or, if it is empty, the update
stack) will contain a \lstinline{case} expression to match against this result.

\section{\emph{Francesca}~-- RPython Usage \& Notes} %% RPython

In this section, we discuss a number of intricacies that require careful
consideration during implementation of the STG machine before describing
how we have leveraged the \lstinline{jitdriver} in our implementation.

\subsection{RPython Customizations}

\paragraph{Unique Stack Implementations.}\label{sec:typeinf}
As previously state, RPython attempts exhaustive type inference for structures
during compilation. This approach has forced us into using three different
stack implementations, all identical in code; we would like to use a single
stack defined as \lstinline{data/stack.py}, but these authors could not find
a way to circumvent the type inference to allow such reuse.
Indeed, we would suggest that this is one of the largest
hurdles in constructing a project such as this in RPython: there is a non-zero
amount of code duplication required to simply exist in the RPython ecosystem.

Furthermore, the machine description~\cite{spj:stgmachine} describe an
optimization for combining stacks that would suggest that we can combine these
stacks by carefully packaging each different node in a new object to indicate
which of the original stacks it lived on\footnote{There are other mechanisms to
avoid this discussed in the original work, but they require subtle bit
manipulation that would need to be faked in a higher-level implementation.}. 
Unfortunately, this would add one further level of indirection for the compiler
and further obfuscate our implementation, and it is unclear if we could gain
speed for the effort.  Further experimentation should be done, here, however.

\paragraph{The Funny Case of Laziness.}
While standard lazy languages provide forcing at primitive operations, the
STG machine departs from this explicit approach. Instead, a primitive operation
assumes that its arguments are pre-evaluated, and this became troublesome when
testing them. To further elaborate this point, observe a standard definition
of addition in the STG language:
\begin{code}[language=Haskell]
+ = \ {} \\n {n,m} -> case n of
                        n' -> case m of
                                m' -> #+ n' m'
\end{code}
This snippet is interpreted as: 
\begin{center}
\emph{
``Plus is a lambda with no free variables that
is should not be updated. When applied, it will \lstinline{case} on
\lstinline{n} and then \lstinline{m}, using the forcing position of
\lstinline{case} to ensure that both inputs are evaluated before using the
primitive addition operator \lstinline{\#+} on the result.''
}
\end{center}
While not particularly tied to the implementation in RPython, we found this
insight invaluable during implementation and testing and so we have included
it here. This manifests with the following dispatch for addition in our
implementation:
\begin{code}
elif expr_type is ast.ast.PrimOp:
  if cexp.oper == "+":
    lookups = vals(code.env,
                   global_env, 
                   cexp.atoms)
    x1      = lookups[0]
    x2      = lookups[1]
    assert isinstance(x1,ast.ast.ValAST)
    assert isinstance(x2,ast.ast.ValAST)
    res     = x1.value + x2.value
    config.code = op.ReturnInt(res) 
\end{code}
We first ensure that we are performing addition and then look up the operators
with \lstinline{vals} (which uses the local and global environments for variable
lookup). Next, we ensure that each of the resulting lookups are valid Value
nodes, extract their values, and then construct a \lstinline{ReturnIn} operation
code for the configuration.

\subsection{The RPython JIT Driver.}

\begin{code}
from rpython.rlib.jit import JitDriver, purefunction

def get_location(code):
    return "%s" % ( code )

jitdriver = JitDriver(greens=['code']
                     , reds='auto'
                     , get_printable_location=get_location)

def terminateHuh(config):
  return (((config.code.op == op.ReturnConOp) or 
           (config.code.op == op.ReturnIntOp)) 
          and config.ret_stack.empty()
          and config.upd_stack.empty())

def loop(config):

  while not terminateHuh(config):
    jitdriver.jit_merge_point( code       = config.code)
    config = config.code.step(config)
  return config
\end{code}

\section{\emph{Neptune} n\'{e}e \emph{Fortune}~-- Conclusion} %% Conclusion / Future Work
